---
title: "Using FARScoursera"
author: "Jake Markman"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
## Welcome to FARScoursera!  

This package is submitted as the final assignment for the Building R Packages class
in the Advanced R Programming specialization offered on Coursera.  

The functions in the package were authored by the course professors and
administrators, while documentation, raw data, and associated package building was provided
by the students.  

FARScoursera itself, as a package, is quite simple (as building i it is a teaching tool).  It provides a suite 
of functions for interacting with the [FARS](https://www.nhtsa.gov/research-data/fatality-analysis-reporting-system-fars) 
database from the [NHTSA](https://www.nhtsa.gov/).  See the documentation in those sites for information as to the data in 
the files themselves.  

The package includes three years of the FARS data in the `extdata` directory, as well as functions to help
load and analyze the data.  

This vignette has sections for each of the five exported functions, as well as one each
for copying the FARS files to the working directory and removing them

### Contents  

1.  [Copy FARS data to working directory](#copy)  
2.  [fars_read](#fr)  
3.  [make_filename](#mknm)  
4.  [fars_read_years](#fry)  
5.  [fars_summarize_years](#fsy)  
6.  [fars_map_state](#fms)  
7.  [Remove FARS data from the working directory](#delete)

### <a name="copy"></a> Copy FARS data to working directory  

The functions below assume that the FARS data is already in your working directory.  You can 
download the data yourself from [FARS](https://www.nhtsa.gov/research-data/fatality-analysis-reporting-system-fars),
however the three years 2013-2015 are included with the package.  To get them into the working directory try the following:  

```{r}
# we need the package, of course:
library(FARScoursera)

wd_path <- getwd()
data_path <- system.file('extdata', package = 'FARScoursera')
years <- 2013:2015

for(y in years) {
    file.copy(paste0(data_path,'/',make_filename(y)),wd_path, overwrite = TRUE)
}
```  
Now we are ready to proceed.  Be sure to remove the files from the working directory if necessary.  

### <a name="fr"></a> fars_read  

`fars_read` is a function that makes it easy to load the FARS data in as a tbl_df.  It takes a single
argument, the filename, as a string.  If the filename does not exist, it throws an error.  The 
`make_filename` function is handy for easily referring to the correct file.  


```{r error=TRUE}

fars2013 <- fars_read(make_filename(2013))

fars2013

#as an example for an error:

fars_read('nofilehere')

```

### <a name="mknm"></a> make_filename  

`make_filename` is a function for returning the correct name of a file for a given year. 
It takes a single input, year, the four-digit year of the file (which can be either a number or a string). 
The function will return the filename, regardless of whether the file exists.  Correct filenames are of the 
form `accident_YYYY.csv.bz2`, where `YYYY` is the four-digit year.

If the input cannot be coerced to an integer, a string is still returned, although a warning is shown.  

```{r error=TRUE}

make_filename(2013)

# no check on whether the file exists
make_filename(12345)

# input not coerceable to string
make_filename('obvious_mistake')
```

### <a name="fry"></a> fars_read_years  

`fars_read_years` takes a single input, a vector of `years`, and returns a list of dplyr tibbles containting the
FARS data for that year.  The tibble contians only the month of the accident and the year in the
format of the `years` vector.  This is primarily a helper function for
`fars_summarize_years` as much of the data contained in the files is lost.

Any component of `years` which does not correspond to a valid filename will produce a warning, but a list
will still be returned, with NULL in position for each invalid component.  

```{r error = TRUE}

list_years <- fars_read_years(2013:2015)
list_years[[1]]

# an example of a bad entry in years
list_years <- fars_read_years(c('2013','a','2014'))
list_years[[2]]
```

### <a name="fsy"></a> fars_summarize_years  

`fars_summarize_years` is a quick function for exploring accident trends over time. 
This function returns a dplyr tibble with the month in rows and the year in columns.
The entries are the total number of accidents in that year / month.  

`fars_summarize_years` takes a single parameter, `years`, a vector of years as in 
`fars_read_years`.  

If any entry in `years` does not result in a valid file, what is returned will be changed--
the tables are still returned as in `fars_read_years`, but the final combination is not performed.  

```{r errpr = TRUE}

fars_summarize_years(2013:2015)

# bad entry in a single year
fars_read_years(c('2013','a','2014'))

# no good years provided at all
fars_read_years(13:15)

```


### <a name="fms"></a> fars_map_states  

`fars_map_state` is a function which graphically summarizes the data by plotting 
the locations of individual accidents on a map of the given state for the given year.

The function takes two inputs, `state.num` and `year`.  

*   `state.num` is a A scalar of a type that can be coerced to an integer of the integer identifier of the state.
A list of the state identifying integers for the FARS data can be found [here](https://crashstats.nhtsa.dot.gov/#/DocumentTypeList/4)  
*   `year` is a four-digit year identifier of a type that can be coerced to integer as in the other functions in this package.  

If either `state.num` or `year` do not refer to valid states or years, respectively, in the data, an error is thrown.  


```{r error = TRUE}

fars_map_state(1,2015)

fars_map_state(6,2013)

# an example of a bad state code
fars_map_state(99, 2013)

# an example of an unavailable year
fars_map_state(1, 2019)

```

### <a name="delete"></a> Remove FARS data from the working directory  

As a final step, we will remove the files we copied into the working directory:  

```{r}
wd_path <- getwd()
years <- 2013:2015

for(y in years) {
    file.remove(paste0(wd_path,'/',make_filename(y)))
}

```
